---
title: "Coding"
author: "Thomas O'Leary"
date: "8/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to **if else** statements

```{r if else statement}
x <- 8
if (x <= 10){
  print("x is less than 10")
} else if (x == 10){
  print("x is 10")
} else {
  print("x is greater than 10")
}
```

### Challenge:
Use an if() statement to print a suitable message reporting whether there are 
any records from the year 2002 in the gapminder dataset. Now do the same for 
the year 2012.

```{r challenge 1}
gapminder <- read.csv("/Users/tsoleary/R/play/r_lessons/gapminder_data.csv")

head (gapminder)

if (sum(gapminder$year == 2002) > 0) {
  num <- sum(gapminder$year == 2002)
  print(paste("2002 is present", num, "times"), quote = T)
} else {
  print("2002 is not present", quote = FALSE)
}

if (sum(gapminder$year == 2012) > 0) {
  print("2012 is present", quote = FALSE)
} else {
  print("2012 is not present", quote = FALSE)
}

```

## Intro to **loops**

### for loops
```{r loops}
for (i in 1:10){
  print(i)
}
```
```{r}
for (i in 1:10){
  print(i)
}
```



```{r}
ali <- 1
while (ali <= 10){
  for (comp in c("a", "b", "c")){
    print(paste(ali, comp))
  }
  ali <- ali + 1
}
```

### while loops
while a certian condition is true it will perform a certain action. It is sorta
as if for loops and if statements had a baby. It is great if you don't know 
how long something is going to take. For example, if you want to do something
until a population reaches a certain value, say one million. You may not know
how long that is going to take but you want it to stop once it reaches there.

```{r}
# go only while z is greater than 0.1 so when it is less than 0.1 it will stop
z <- 1
while (z > 0.1) {
  z <- runif(1) # unif is the uniform distribution and r is for random
  print(z)
}
```


## Intro to functions

Let's try to make a function that converts fahrenheit to kelvin

```{r}
source("/Users/tsoleary/R/play/r_lessons/functions.R")

f_to_k2(32)

kelvin_to_fahr <- function(temp_k){
  kelvin <- (temp_k - 273.15) * 9/5 + 32
  return(kelvin)
}

kelvin_to_fahr(300)
```

# Final Challenge of the day

Create a function that calculates a gdp of a population in the gapminder data

```{r}
calc_gdp <- function(row){
  gapminder[row, "gdpPercap"] * gapminder[row, "pop"]
}

calc_gdp(1)

# so to get crazy and get all of them you can run a for loop
gdps <- NULL
for (i in 1:nrow(gapminder)){
  gdp <- calc_gdp(i)
  gdps <- c(gdps, gdp)
}

head(gdps) # to see what the first 6 rows look like

# so if you wanted to add that to the gapminder dataframe you could do something 
# like this

gapminder$gdp_total <- gdps

head(gapminder) # new column should be present

```

