---
title: Baseball statistics
author: "TS O'Leary"
output:
  rmarkdown::html_document:
    theme: lumen
    number_sections: false
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
# Don't show echos, warnings, or messages
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Load packages
require(tidyverse)
require(Lahman)
require(plotly)

# Load data, filter to modern era AL & NL teams, & calc other stats ------------
teams <- Teams %>% 
  filter(yearID >= 1901 & 
           lgID %in% c("AL", "NL")) %>%
  group_by(yearID, teamID) %>%
  mutate(TB = H + X2B + 2 * X3B + 3 * HR,
         WinPct = W/G,
         RpG = R/G,
         RunDiff = R - RA,
         HRpG = HR/G,
         TBpG = TB/G,
         SOpG = SO/G,
         SFpG = SF/G,
         `SO:BB` = SO/BB,
         KpIP = SOA/(IPouts/3),
         `HR:SO` = HR/SO,
         BABIP = (H - HR) / (AB - HR - SO + SF),
         WHIP = 3 * (H + BB)/IPouts)

# Get batting stats
Batting <- battingStats(data = Batting, 
                        idvars = c("playerID", 
                                   "yearID", 
                                   "stint", 
                                   "teamID", 
                                   "lgID"), 
                        cbind = TRUE)
```


# **All teams**

## Runs scored per game

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(RpG = mean(RpG)) %>%
  ggplot(aes(x = yearID, y = RpG)) +
  geom_line() +
  theme_classic()

ggplotly(p)
```

## Home runs per game

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(HRpG = mean(HRpG)) %>%
  ggplot(aes(x = yearID, y = HRpG)) +
  geom_line() +
  theme_classic()

ggplotly(p)
```

## Strikeouts per game

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(SOpG = mean(SOpG, na.rm = TRUE)) %>%
  ggplot(aes(x = yearID, y = SOpG)) +
  geom_line() +
  theme_classic()

ggplotly(p)
```

## Home Run to Strikeout ratio

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(`HR:SO` = mean(`HR:SO`, na.rm = TRUE)) %>%
  ggplot(aes(x = yearID, y = `HR:SO`)) +
  geom_line() +
  theme_classic()

ggplotly(p)
```


## Ks per IP

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(KpIP = mean(KpIP, na.rm = TRUE)) %>%
  ggplot(aes(x = yearID, y = KpIP)) +
  geom_line() +
  theme_classic()

ggplotly(p)
```

## Complete games

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(CG = sum(CG)) %>%
  ggplot(aes(x = yearID, y = CG)) +
  geom_line() +
  theme_classic()

ggplotly(p) 
```

## ERA

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(ERA = mean(ERA)) %>%
  ggplot(aes(x = yearID, y = ERA)) +
  geom_line() +
  theme_classic()

ggplotly(p) 
```

## Total bases per game

```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(TBpG = mean(TBpG)) %>%
  ggplot(aes(x = yearID, y = TBpG)) +
  geom_line() +
  theme_classic()

ggplotly(p) 
```

## Batting average on balls in play
```{r}
p <- teams %>%
  group_by(yearID) %>%
  summarise(BABIP = mean(BABIP)) %>%
  ggplot(aes(x = yearID, y = BABIP)) +
  geom_line() +
  theme_classic()

ggplotly(p) 
```




# **World Series Champions**

## Runs scored per game

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = RpG, color = teamID)) +
  geom_point() +
  theme_classic()

ggplotly(p)
```

## Home runs per game

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = HRpG, color = teamID)) +
  geom_point() +
  theme_classic()

ggplotly(p)
```

## Strikeouts per game

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = SOpG, color = teamID)) +
  geom_point() +
  theme_classic()

ggplotly(p)
```

## Run differential

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = RunDiff, color = teamID)) +
  theme(legend.position = "none") +
  geom_point() +
  theme_classic()

ggplotly(p) 
```

## Win percentage

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = WinPct, color = teamID)) +
  theme(legend.position = "none") +
  geom_point() +
  theme_classic()

ggplotly(p) 
```

## Ks per IP

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = KpIP, color = teamID)) +
  geom_point() +
  theme_classic()

ggplotly(p)
```

## Complete games

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = CG, color = teamID)) +
  theme(legend.position = "none") +
  geom_point() +
  theme_classic()

ggplotly(p) 
```

## ERA

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = ERA, color = teamID)) +
  theme(legend.position = "none") +
  geom_point() +
  theme_classic()

ggplotly(p) 
```

## Total bases per game

```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = TBpG, color = teamID)) +
  theme(legend.position = "none") +
  geom_point() +
  theme_classic()

ggplotly(p) 
```


## Batting average on balls in play
```{r}
p <- teams %>%
  filter(WSWin == "Y") %>%
  ggplot(aes(x = yearID, y = BABIP, color = teamID)) +
  geom_point() +
  theme_classic()

ggplotly(p) 
```