---
title: "Single-species models in discrete time"
subtitle: "Problem Set 1"
author: "Thomas O'Leary"
date: "September 11, 2019"
geometry: "left = 2cm, right = 2cm, top = 1.5cm, bottom = 1.5cm"
output: 
  pdf_document:
    fig_width: 6
    fig_height: 3.75
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\section{1. Discrete-time logistic growth model: Increasing r}
Modify the following code to determine the long-term behavior of the discrete-time logistic growth model. In the example below, the population movestowards an equilbrium point. How does the long-term behavior change as you increase $r$ slowly.

```{r, echo = TRUE, eval = TRUE}
plot_log_growth <- function(r, max_time = 50, K = 1, FUN = points, 
                            color = "black"){
  # initialize vector
  N <- vector('numeric',length = max_time)
  N[1] <- 0.1
  # iterate model
  for (t in 1:max_time){
    N[t+1] <- N[t] + r*N[t]*(1 - N[t]/K)
  }
  # plot N over time
  FUN(N, type = 'l', ylim = c(0, 1.25), col = color, 
      main = "Logistic growth at increasing values of r", xlab = "time (t)",
      ylab = "N (population size)")
}
# plot a bunch of different growth curves at increasing values of r
plot_log_growth(r = 0.4, FUN = plot)
plot_log_growth(r = 1, color = "red")
plot_log_growth(r = 1.4, color = "orange")
plot_log_growth(r = 1.8, color = "yellow")
plot_log_growth(r = 2.0, color = "green")
plot_log_growth(r = 2.5, color = "blue")
```


As $r$ increases the amount of fluctuation at the equilibrium point increases. So at $r = 0.4$ $N$ gradually increases to 1.0 and then stays at equilbrium. At higher $r$ values the population reaches quilibrium faster but at values close to 2.0 the equilibrium begins to oscillate above and below a N of 1.0. 


\section{2. Plot the final ten years of N as a function of increasing values of $r$ }
To show the results in question 1 more clearly, write R code to build a plot of the equilibrium value versus the value for $r$. To do this, run the model for 50 years, and only plot the last 10 years. If the solution is an equilibrium point, all 10 years should be the same values.

```{r, echo = TRUE, eval = TRUE}
max_time <- 50
K <- 1

n_vals <- NULL
r_vals <- NULL
for (r in seq(from = 0.4, to = 3, by = .1)){
  # initialize vector
  N <- vector('numeric',length = max_time)
  N[1] <- 0.1
  # iterate model
  for (t in 1:max_time){
    N[t+1] <- N[t] + r*N[t]*(1 - N[t]/K)
  }
  # save only the final 10 years and the corresponding r values
  n_vals <- c(n_vals, N[(max_time-9):max_time])
  r_vals <- c(r_vals, rep(r, 10))
}

vals <- as.data.frame(cbind(r_vals, n_vals))

# load ggplot2 library to run ggplot()
library(ggplot2)

ggplot(vals, aes(x = r_vals, y = n_vals)) +
  geom_jitter(mapping = aes(x = r_vals, y = n_vals), fill = "grey", 
        alpha = 0.5, size = 3, pch = 21, color = "black", width = 0.02) +
  theme_classic() + 
  labs(x = "r value", y = "N (population size)", 
       title = "Population size at \"equilibrium\" as a function of r")
```




\section{3. Fisheries Example}

In order to increase fisheries yields, models of fish populations have been used. Two common models are Beverton-Holt and the Ricker. Here we will use the Beverton-Holt model with parameters *a* and *b*:

$$
N(t+1) = \frac{a N(t)}{1 + b N(t)} 
$$

```{r cod_pop, echo = TRUE, eval = TRUE, fig.cap = "Left: Canadian Atlantic cod population (in tonnes) over time and Right: Phase plane plot of the Atlantic cod population", fig.pos = 'b'}
cod_pop <- c(1450, 1420, 1050, 2643, 1060, 1080, 1410, 1150, 740, 175, 43, 10, 
             12, 15, 16, 16, 28, 30, 32, 23, 12, 19, 27)
years <- c(1983, 1984, 1985, 1986, 1987, 1988, 1989, 1990, 1991, 1992, 
           1993, 1994, 1995, 1996, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 
           2004, 2005)

par(mfrow = c(1, 2))
plot(years, cod_pop, ylim = c(0, 3000), las = 1, ylab = 'Population (tonnes)', 
     xlab='Time (years)', pch = 16)
plot(cod_pop[1:(length(cod_pop) - 1)], cod_pop[2:length(cod_pop)], 
     ylab = 'N[t+1]', xlab = 'N[t]', las = 1, pch = 16)
mtext(text = 'Atlantic cod population - Canada', side = 3, line = -3, 
      outer = TRUE)
```

In order to increase fisheries yields, models of fish populations have been used. Two common models are Beverton-Holt and the Ricker. Here we will use the Beverton-Holt model with parameters *a* and *b*:

$$
N(t+1) = \frac{a N(t)}{1 + b N(t)} 
$$

\subsection{A. What are the parameter values (*a* and *b*) that best fit the data in phase plane diagram above?}

```{r parameter est, echo = TRUE, eval = TRUE}
# use a curve fitting package
# exponential equation to

# bev_holt <- function(n, a, b) {
#   (a * n) / (1 + (b * n))
# }
# 
# fitPool <- nls(y ~ bev_holt(n, a, b), data = newDF,
#                start = list(Topt=25,Tmax=35,Ymax=2.5))
# 
# nls(ratio ~ SSasymp(week, yf, y0, log_k), data = .))
# 
# formula = y ~ SSasymp(n, a, b)
# 
# 
# 
# 
# ggplot(dat, aes(x = week, y = ratio)) +
#   FUN(mapping = aes(x = week, y = ratio, fill = leg),
#       alpha = 0.5, size = 3, pch = 21,  color = "black", width = 0.05) +
#   geom_smooth(method = "nls", formula = y ~ SSasymp(x, yf, y0, log_k),
#               data = dat,
#               se = FALSE,
#               aes(color = leg), show.legend = FALSE)



```



<!-- #### 3b) For the best fit model, what is the optimal level of harvesting (*h*) that maximizes yield and prevents the population from going below 100 tonnes over a 50 year period? Use the modified Beverton-Holt model with a harvesting rate of *h*: -->

<!-- $$ -->
<!-- N(t+1) = (1-h) \frac{a N(t)}{1 + b N(t)}  -->
<!-- $$ -->




