---
title: A Skeptic's Analysis of Scientific Writing
author: "Team Ice Dragons"
output:
  rmarkdown::html_document:
    theme: lumen
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
# don't show echos, warnings or messages
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE,
                      comment = NA)



# Load packages
require(tidyverse)
#require(stringi)

# Load data
df <- read_csv("https://raw.githubusercontent.com/StefanoAllesina/QuEST_workshop/main/data/plos_compbio.csv")
```

Analysis of data for [Dr. Stefano Allesina's workshop: "A Skeptic's Guide to Scientific Writing."](https://stefanoallesina.github.io/QuEST_workshop/)




I think for the most part I am going to focus only on **Articles**.

Dr. Allesina visualized some of the data in a time-series format [here](https://stefanoallesina.github.io/QuEST_workshop/#taking-a-peek). I think I am just going to look at the data regardless of time.


# **Distributions**

## Citations

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot(aes(x = log(num_citations + 1))) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(log(df$num_citations + 1)), linetype = 2) +
  theme_classic()
```

## Views

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot(aes(x = log(num_views + 1))) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(log(df$num_views + 1)), linetype = 2) +
  theme_classic()
```

## Authors

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot(aes(x = num_authors)) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(df$num_authors), linetype = 2) +
  theme_classic()
```


## Figures

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot(aes(x = num_figures)) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(df$num_figures), 
             linetype = 2) +
  annotate("text", 
           label = paste("median =", median(df$num_figures)), 
           y = 1000, 
           x = median(df$num_figures)*2) +
  theme_classic()
```

## Words in the abstract

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot(aes(x = num_words_abstract)) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(df$num_words_abstract), 
             linetype = 2) +
  theme_classic()
```

## Words in the title 

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot(aes(x = num_words_title)) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(df$num_words_title), 
             linetype = 2) +
  theme_classic()
```

## Words in the dictionary in the abstract

Filtered out "0" values.

```{r}
df %>%
  filter(document_type == "Article",
         prop_words_abs > 0) %>%
  ggplot(aes(x = prop_words_abs)) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(df$prop_words_abs), 
             linetype = 2) +
  theme_classic()
```

## Simple words in the abstract

Filtered out "0" values.

```{r}
df %>%
  filter(document_type == "Article",
         prop_simple_words_abs > 0) %>%
  ggplot(aes(x = prop_simple_words_abs)) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(df$prop_simple_words_abs), 
             linetype = 2) +
  theme_classic()
```

## References

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot(aes(x = num_references)) +
  geom_histogram(fill = "grey90", color = "grey50") +
  geom_vline(xintercept = median(df$num_references), 
             linetype = 2) +
  theme_classic()
```

# **Citations**

## Authors

```{r}
df %>%
  filter(document_type == "Article") %>%
  mutate(num_authors = case_when(num_authors < 9 ~ as.character(num_authors),
                                 num_authors >= 9 ~ "9 or more")) %>%
  ggplot(aes(x = num_authors, y = log(num_citations + 1))) +
  geom_violin(fill = "azure2") +
  geom_boxplot(width = 0.15) +
  theme_classic()
```

## Countries

```{r}
df %>%
  filter(document_type == "Article") %>%
  mutate(num_countries = case_when(num_countries < 9 ~ as.character(num_countries),
                                   num_countries >= 9 ~ "9 or more")) %>%
  ggplot(aes(x = num_countries, y = log(num_citations + 1))) +
  geom_violin(fill = "azure2") +
  geom_boxplot(width = 0.15) +
  theme_classic()
```

## Equations

```{r}
df %>%
  filter(document_type == "Article") %>%
  mutate(num_countries = case_when(num_countries < 9 ~ as.character(num_countries),
                                   num_countries >= 9 ~ "9 or more")) %>%
  ggplot(aes(x = num_countries, y = log(num_citations + 1))) +
  geom_violin(fill = "azure2") +
  geom_boxplot(width = 0.15) +
  theme_classic()
```

# **Correlation of views and citations**

Melissa brought up an interesting question as to how related the number of veiws and citations are. Stefano mentioned that for some types of documents they may not be closely related at all.

```{r}
# Plot num_views vs. num_citations
df %>%
  ggplot() +
  geom_point(aes(x = log(num_views + 1), 
                 y = log(num_citations + 1)),
             shape = 21, 
             color = "grey50", 
             alpha = 0.8)  +
  facet_wrap(~ document_type) +
  theme_classic()
```

```{r}
# Plot num_views vs. num_citations
df %>%
  filter(document_type %in% c("Article", "Editorial", "Review")) %>%
  ggplot() +
  geom_point(aes(x = log(num_views + 1), 
                 y = log(num_citations + 1)),
             shape = 21, 
             color = "grey50", 
             alpha = 0.8)  +
  facet_wrap(~ document_type) +
  theme_classic()
```

# **How does the number of words in the title affect the number of citations?**

## Visualize

```{r}
df %>%
  filter(document_type == "Article") %>%
  ggplot() +
  geom_point(aes(x = num_words_title, 
                 y = log(num_citations + 1)),
             shape = 21, 
             color = "grey50", 
             alpha = 0.8)  +
  theme_classic()
```

## GLM

```{r}
dat <- df %>%
  filter(document_type == "Article")

fit <- glm(log(num_citations + 1) ~ num_words_title,
           data = dat)

summary(fit)

print("p-value")
coef(summary(fit))[,4]
```

# **Titles with ":"**

```{r}
dat <- df %>%
  filter(document_type == "Article") %>%
  mutate(colon = str_detect(title, ":"))

print("Number of titles with a colon")
sum(dat$colon)

print("Proportion of titles with a colon")
sum(dat$colon)/nrow(dat)
```

```{r}
dat %>%
  ggplot(aes(x = colon, y = log(num_citations + 1))) +
  geom_violin(fill = "azure2") +
  geom_boxplot(width = 0.15) +
  theme_classic()
```

## Unique characters in the title

```{r}
all_titles <- paste(dat$title, collapse = " ")

str_extract_all(all_titles, boundary("character"))%>% 
  unlist() %>% 
  as_tibble() %>% 
  count(value)
```

